Models

### 2021-04-14 analysis/05_model_metrics.Rmd
Clean up plotting, update model output processing to reflect lmerTest output

```{r}
library(tidyverse)
#library(moderndive)
library(lme4)
library(broom.mixed)
```

@@

Visualize distributions of response
```{r}
metrics %>%
  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
  ggplot() +
  geom_histogram(aes(x = logvalue),
                 bins = 60) +
  facet_wrap(~metric, scales = "free") +
  labs(title = "Distribution of Log values")

metrics %>%
  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "SES_FDis", "SES_FDiv", "SES_FEve", "SES_FRic")) %>%
  ggplot() +
  geom_histogram(aes(x = value),
                 bins = 60) +
  facet_wrap(~metric, scales = "free") +
  labs(title = "Distribution of raw values")
```

@@

SES_rand_slope <- model_data %>%
  #filter(protected_area == "FALSE") %>%
  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis")) %>%
  filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           test = map2(model, null_model, anova),
           test_res = map(test, tidy),
           coef = map(model, tidy)
           ) %>% 
  unnest(cols = c(test_res)) %>%
  select(metric, model, coef, p.value) %>%
  filter(!is.na(p.value)) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

metric_rand_slope <- model_data %>%
  #filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
  filter(metric %in%  c("FRic", "FEve", "FDiv", "S")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x, REML = FALSE)),
           null_model = map(data, ~lmer(logvalue ~ (year_scaled|study_id/rarefyID), 
                                 data = .x, REML = FALSE)),
           test = map2(model, null_model, anova),
           test_res = map(test, tidy),
           coef = map(model, tidy)
           ) %>% 
  unnest(cols = c(test_res)) %>%
  select(metric, model, coef, p.value) %>%
  filter(!is.na(p.value)) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

@@

Kitchen sink model
```{r}
all_model <- model_data %>%
  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + taxa + realm + (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           null_model = map(data, ~lmer(value ~ taxa + realm + (year_scaled|study_id/rarefyID),
                                 data = .x)),
           test = map2(model, null_model, anova),
           test_res = map(test, tidy),
           coef = map(model, tidy)
           ) %>%
  unnest(cols = c(test_res)) %>%
  select(metric, model, coef, p.value) %>%
  filter(!is.na(p.value)) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))
```

Reproduce Dornelas figure - all data points & regressions w/overall mean
```{r}
#get climate classification
metadata <- read_csv(system.file("extdata", "biotime/biotime_metadata.csv", package = "biodivTS"))

@@

plot_metric <- function(metric_name, ylabel, data, model_coef){

@@

  trend_plot <- data %>%
    left_join(pred, by = "year_scaled") %>%
    #left_join(metadata %>% select(study_id, climate)) %>%
    ggplot(aes(x = year_scaled, y = logvalue)) +
    # geom_smooth(aes(color = climate, group = rarefyID), method = "glm", se = FALSE) +
    # geom_point(aes(color = climate, group = rarefyID)) +
    geom_point(aes(group = rarefyID) ,color = "grey", size = 0.30) +
    geom_smooth(aes(color = climate_mod, group = rarefyID), method = "glm", se = FALSE, size = 0.5) +
    #geom_smooth(color = "black", method = "glm", se = FALSE) +
    geom_line(mapping = aes(y = pred)) +
    theme(legend.position = "none") +
    theme_classic() +
    ylab(ylabel) +
    labs(color = "Climate")

  ggsave(paste0(metric_name, "_trend.png"), trend_plot)
  return(trend_plot)
}

model_coef <- metric_mean_fit_slope %>%
  select(metric, model) %>%
   mutate(coef = map(model, tidy)) %>% 
  select(-model) %>% 
  unnest(cols = c(coef))

@@

plot_map <- tibble(metric_name = c("S_int", "FRic", "FEve", "FDiv", "FDis"), 
           ylabel = c("log(S_int)", "log(FD)", "log(Evenness)", "log(Divergence)", "log(Dispersion)"))
pmap(plot_map, plot_metric, data = metrics, model_coef = model_coef)
```

blup <- pmap_dfr(bind_rows(select(SES_rand_slope, metric, model), select(metric_rand_slope,  metric, model)), get_blup) %>%
  select(-"(Intercept)") %>%
  pivot_wider(names_from = "metric", values_from = "year_scaled")

#pairwise plots using GGally
ggpairs(blup %>% 
          select(-c("study_id", "FDis", "FDiv", "FEve", "FRic")))

blup %>% 
  gatherpairs("SES_FDis", "SES_FDiv", "SES_FEve", "SES_FRic", "S") %>% {
  ggplot(., aes(x = .xvalue, y = .yvalue)) +
      geom_point() + 
      #geom_smooth(method = 'lm') +
      facet_wrap(.xkey ~ .ykey, ncol = length(unique(.$.ykey)), scales = 'free', labeller = label_both) +
      scale_color_brewer(type = 'qual')
  }

blup %>%
  select(-c("FDis", "FDiv", "FEve", "FRic")) %>%
  pivot_longer(-one_of("study_id", "S"), names_to = "metric") %>%
  ggplot(aes(x = S, y = value)) +
  geom_point() +
  facet_wrap(~metric) +
  xlim(-1.3, 1.3) +
  ylim(-0.6, 0.6) +
    theme_classic() +
    geom_hline(yintercept = 0) +
    geom_vline(xintercept = 0) + 
  ylab("Metric Slope") +
  xlab("Richness Slope")

blup %>%
  select(-c("FDis", "FDiv", "FEve", "FRic")) %>%
  pivot_longer(-one_of("study_id", "SES_FRic"), names_to = "metric") %>%
  ggplot(aes(x = SES_FRic, y = value)) +
  geom_point() +
  facet_wrap(~metric) +
  xlim(-.25, .25) +
  ylim(-0.6, 0.6) +
    theme_classic() +
    geom_hline(yintercept = 0) +
    geom_vline(xintercept = 0) + 
  ylab("Metric Slope") +
  xlab("Functional Richness Slope")
```

Clustering 
```{r}
scale2 <- function(x, na.rm = FALSE) (x - mean(x, na.rm = na.rm)) / sd(x, na.rm)

blup_mat <- blup %>%  
  #these studies have NA's for some reason, remove them 
  filter(!study_id %in% c(445, 446, 348, 382)) %>%
  select(study_id, starts_with("SES"), S) %>%
  column_to_rownames("study_id") %>%
  mutate(across(.fns = scale2)) %>%
  as.matrix()

blup_clust <- blup %>% 
  left_join(as.data.frame(cat3) %>%
  rownames_to_column("study_id")) %>%
  left_join(as.data.frame(cat5) %>%
  rownames_to_column("study_id"))
```

>>>>

```{r}
library(tidyverse)
#library(moderndive)
library(lme4)
library(broom.mixed)
library(lmerTest)
```

@@

Visualize distributions of response
```{r}
metrics %>%
  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "Jaccard_hind")) %>%
  ggplot() +
  geom_histogram(aes(x = logvalue),
                 bins = 60) +
  facet_wrap(~metric, scales = "free") +
  labs(title = "Distribution of Log values")

metrics %>%
  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "SES_FDis", "SES_FDiv", "SES_FEve", "SES_FRic", "Jaccard_hind")) %>%
  ggplot() +
  geom_histogram(aes(x = value),
                 bins = 60) +
  facet_wrap(~metric, scales = "free") +
  labs(title = "Distribution of raw values")
```

@@

SES_rand_slope <- model_data %>%
  #filter(protected_area == "FALSE") %>%
  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis", )) %>%
  filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "Jaccard_next")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           # null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
           #                       data = .x)),
           # test = map2(model, null_model, anova),
           # test_res = map(test, tidy),
           coef = map(model, tidy)
           ) %>% 
  select(metric, model, coef) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

metric_rand_slope <- model_data %>%
  #filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
  filter(metric %in%  c("FRic", "FEve", "FDiv", "S")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x, REML = FALSE)),
           coef = map(model, tidy)
           ) %>% 
  select(metric, model, coef) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

@@

Kitchen sink model
```{r}
all_model <- model_data %>%
  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + taxa + realm + (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           null_model = map(data, ~lmer(value ~ taxa + realm + (year_scaled|study_id/rarefyID),
                                 data = .x)),
           test = map2(model, null_model, anova),
           test_res = map(test, tidy),
           coef = map(model, tidy)
           ) %>%
  unnest(cols = c(test_res)) %>%
  select(metric, model, coef, p.value) %>%
  filter(!is.na(p.value)) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))
```
Models with categorical study characteristics as random effects

Taxa:
```{r}
taxa_slope_model <- model_data %>%
  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + taxa + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
                                 data = .x, REML = TRUE)),
           coef = map(model, tidy)) %>%
  select(metric, coef) %>%
  unnest(cols = c(coef)) %>%
  #filter(term == "year_scaled") %>%
  filter(term %in% c("year_scaled", "taxaBirds", "taxaMammals")) %>%
  select(-c(effect, group, statistic))

taxa_slope_model_S <- model_data %>% 
  filter(metric == "S") %>%
  lmer(logvalue ~ year_scaled + taxa + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
                                 data = ., REML = TRUE)
```
Realm:
```{r}
realm_slope_model <- model_data %>%
  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + realm + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
                                 data = .x, REML = TRUE)),
           coef = map(model, tidy)) %>%
  select(metric, coef) %>%
  unnest(cols = c(coef)) %>%
  #filter(term == "year_scaled") %>%
  filter(effect == "fixed") %>%
  select(-c(effect, group, statistic))

realm_slope_model_S <- model_data %>% 
  filter(metric == "S") %>%
  lmer(logvalue ~ year_scaled + realm + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
                                 data = ., REML = TRUE)
```
Climate:
```{r}
climate_slope_model <- model_data %>%
  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + climate + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
                                 data = .x, REML = TRUE)),
           coef = map(model, tidy)) %>%
  select(metric, coef) %>%
  unnest(cols = c(coef)) %>%
  #filter(term == "year_scaled") %>%
  filter(effect == "fixed") %>%
  select(-c(effect, group, statistic))

climate_slope_model_S <- model_data %>% 
  filter(metric == "S") %>%
  lmer(logvalue ~ year_scaled + climate + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
                                 data = ., REML = TRUE)
```

blup <- pmap_dfr(bind_rows(select(SES_rand_slope, metric, model), select(metric_rand_slope,  metric, model)), get_blup) %>%
  rename(intercept = "(Intercept)")

blup_slope <- blup %>%
  pivot_wider(names_from = "metric", values_from = "year_scaled")

#pairwise plots using GGally
ggpairs(blup_slope %>% 
          select(-c("study_id", "FDiv", "FEve", "FRic")))

blup_slope %>% 
  gatherpairs( "SES_FDiv", "SES_FEve", "SES_FRic", "S") %>% {
  ggplot(., aes(x = .xvalue, y = .yvalue)) +
      geom_point() + 
      #geom_smooth(method = 'lm') +
      facet_wrap(.xkey ~ .ykey, ncol = length(unique(.$.ykey)), scales = 'free', labeller = label_both) +
      scale_color_brewer(type = 'qual')
  }

blup_slope %>%
  select(-c("FDiv", "FEve", "FRic")) %>%
  pivot_longer(-one_of("study_id", "S"), names_to = "metric") %>%
  ggplot(aes(x = S, y = value)) +
  geom_point() +
  facet_wrap(~metric) +
  xlim(-1.3, 1.3) +
  ylim(-0.6, 0.6) +
    theme_classic() +
    geom_hline(yintercept = 0) +
    geom_vline(xintercept = 0) + 
  ylab("Metric Slope") +
  xlab("Richness Slope")

blup_slope %>%
  select(-c("FDiv", "FEve", "FRic")) %>%
  pivot_longer(-one_of("study_id", "SES_FRic"), names_to = "metric") %>%
  ggplot(aes(x = SES_FRic, y = value)) +
  geom_point() +
  facet_wrap(~metric) +
  xlim(-.25, .25) +
  ylim(-0.6, 0.6) +
    theme_classic() +
    geom_hline(yintercept = 0) +
    geom_vline(xintercept = 0) + 
  ylab("Metric Slope") +
  xlab("Functional Richness Slope")
```

Clustering 
```{r}
library(clValid)
scale2 <- function(x, na.rm = FALSE) (x - mean(x, na.rm = na.rm)) / sd(x, na.rm)

blup_mat <- blup_slope %>%  
  #these studies have NA's for some reason, remove them 
  filter(!study_id %in% c(445, 446, 348, 382, 312)) %>%
  select(study_id, starts_with("SES"), S) %>%
  column_to_rownames("study_id") %>%
  mutate(across(.fns = scale2)) %>%
  as.matrix()

blup_clust <- blup_slope %>% 
  left_join(as.data.frame(cat3) %>%
  rownames_to_column("study_id")) %>%
  left_join(as.data.frame(cat5) %>%
  rownames_to_column("study_id"))
```
```{r}
blup_clust %>% 
  gatherpairs( "SES_FDiv", "SES_FEve", "SES_FRic", "S") %>% {
  ggplot(., aes(x = .xvalue, y = .yvalue, color = as.factor(cat5))) +
      geom_point() + 
      #geom_smooth(method = 'lm') +
      facet_wrap(.xkey ~ .ykey, ncol = length(unique(.$.ykey)), scales = 'free', labeller = label_both) +
      scale_color_brewer(type = 'qual')
  }
```

##Figures
Reproduce Dornelas figure - all data points & regressions w/overall mean
```{r}
#get climate classification
metadata <- read_csv(system.file("extdata", "biotime/biotime_metadata.csv", package = "biodivTS"))

@@

plot_metric <- function(metric_name, ylabel, yval, data, model_coef){

@@

  trend_plot <- data %>%
    left_join(pred, by = "year_scaled") %>%
    #left_join(metadata %>% select(study_id, climate)) %>%
    ggplot(aes(x = year, y = !!sym(yval))) +
    # geom_smooth(aes(color = climate, group = rarefyID), method = "glm", se = FALSE) +
    # geom_point(aes(color = climate, group = rarefyID)) +
    geom_point(aes(group = rarefyID) ,color = "grey", size = 0.50) +
    geom_smooth(aes(color = climate, group = study_id), method = "glm", se = FALSE, size = 0.7) +
    #geom_smooth(color = "black", method = "glm", se = FALSE) +
    geom_line(mapping = aes(y = pred)) +
    theme(legend.position = "none") +
    theme_classic() +
    ylab(ylabel) +
    labs(color = "Climate") +
    scale_fill_jcolors(palette = "pal7")

  ggsave(paste0(here::here("figures"), "/",metric_name, "_trend.png"), trend_plot)
  return(trend_plot)
}

model_coef <- bind_rows(SES_rand_slope, metric_rand_slope) %>%
  select(metric, model) %>%
   mutate(coef = map(model, tidy)) %>% 
  select(-model) %>% 
  unnest(cols = c(coef))

@@

plot_map <- tibble(metric_name = c("S", "SES_FRic", "Jaccard_base"), 
           ylabel = c("log(S)", "Functional Richness", "Jaccard"),
           yval = c("logvalue", "value", "value"))
m_plots <- pmap(plot_map, plot_metric, data = model_data, model_coef = model_coef)

plot_join <- plot_grid(plotlist=m_plots, labels = c("A", "B", "C"), nrow = 1)
```
